---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Do you notice anything different with this read_csv()? Particularly in the file name?

**A1: It seems to be taking the data from a real website called earthquake.usgs.gov**

------------------------------------------------------------------------

#### **Q2** How many records there are there in this data set? What do they mean and what useful information we can gather from it, looking at the columns?

**A2: There is 9,774 entries in this data set. Every single entry is a different earthquake. We can get the time, exact location, depth, magnitude and more from the data set.**

------------------------------------------------------------------------

#### **Q3** How do I reorganize this list to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}
earthquakes |>
  arrange(desc(depth))
  
```

**A3: The deepest earthquake in the data set has a depth of 669.9820 and a magnitude of 4.20**

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}
earthquakes |>
  filter(mag > 6)

```

**A4: It says there are 12 rows when I do this, but the second page is empty. I'm not sure if this is a problem with my computer of if there is two empty rows that were somehow selected. So either 12 or 10.**

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}
earthquakes |>
  filter(mag > 6 & depth < 20)

```

**A5: There is just five entries that support this criteria.**

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}
earthquakes |>
  filter(mag > 6 | depth < 20)

```

**A6: There is 7,412 earthquakes that fit either of these bounds.**

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}
earthquakes |>
  filter(str_detect(place, 'Alaska'))
```

**A7: You should look at the column "place."** **In the data 3,428 of the earthquakes occurred in Alaska.**

------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}
earthquakes |>
  group_by(type) |>
  summarise(
    count_types = n()
  ) |>
  arrange(desc(count_types))

```

**A8: The first two most common types of tremors are earthquake and quarry blast.**

------------------------------------------------------------------------

#### **Q9** How do I calculate the average depth of each type of tremor? What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}
earthquakes |>
  group_by(type) |>
  summarise(
    avg_depth = mean(depth)
  )

```

**A9: Its not very surprising that explosions and quarry blasts occurred above ground, causing the depth to be negative. Ice quakes occurred at surface level, which was interesting. The most unusual or interesting thing though I would say is the other events. What constitutes an other event? Is it a crowd at a sporting event or concert?**

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}
earthquakes |>
  mutate(hour = hour(time)) |>
  group_by(hour)

```

**A10: See above for answer**

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column making that distinction?

```{r}
earthquakes <- earthquakes |>
  mutate(serious = case_when(
    mag > 3 ~ "Serious",
    .default = "Not Serious"
  ))

```

**A11: See above for answer**

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
earthquakes |>
  mutate(hour = hour(time)) |>
  filter(serious == 'Serious' & type == 'earthquake') |>
  group_by(hour) |>
  summarise(
    count_hour = n()
  ) |>
  arrange(desc(count_hour))

```

**A12**: **The 2nd hour of the day has the most earthquakes and for some reason I once again cannot access the last page with four rows.**

------------------------------------------------------------------------

#### EXTRA Q What other questions are you interested in and how would you ask them in R?

```{r}

#Is there a correspondence between seriousness of earthquakes and location?
```

```{r}

#How often was the depth above ground for events that caused earthquakes?
```

```{r}

#What times of day (on average) have the most serious earthquakes?
```
